---
title: Reproducible Research - Peer Assessment 1
author: "by Donatella Biancone"
output:
  html_document
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(scales)
library(Hmisc)
```

## Loading and preprocessing the data
##### 1. Load the data (i.e. read.csv())
```{r, results='markup', warning=TRUE, message=TRUE}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
activityData <- read.csv('activity.csv')
```
##### 2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
activityData$date <- as.Date(activityData$date, "%Y-%m-%d")
```

-----

## What is mean total number of steps taken per day?
```{r}
totalSteps <- tapply(activityData$steps, activityData$date, FUN=sum, na.rm=TRUE)
```

##### 1. Make a histogram of the total number of steps taken each day
```{r}
qplot(totalSteps, main = "Total Number of Steps Taken Each Day", binwidth=1000, xlab="total number of steps")
```

##### 2. Calculate and report the mean and median total number of steps taken per day
```{r}
totalStepsMean <- mean(totalSteps)
totalStepsMedian <- median(totalSteps)
```

-----

## What is the average daily activity pattern?
```{r}
avgSteps <- aggregate(x=list(meanSteps=activityData$steps), by=list(interval=activityData$interval), FUN=mean, na.rm=TRUE)
```

##### 1. Make a time series plot
```{r}
ggplot(data=avgSteps, title = "Time Series Plot of the 5-minute Interval", aes(x=interval, y=meanSteps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken") 
```

##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
avgSteps[which.max(avgSteps$steps),]
```

----

## Imputing missing values
##### 1. Calculate and report the total number of missing values in the dataset 
```{r}
numMissingValues <- length(which(is.na(activityData$steps)))
```

##### 2. Devise a strategy for filling in all of the missing values in the dataset.
We fill NA values in data column using the mean for that 5-minute interval

##### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activityDataFilled <- activityData
activityDataFilled$steps <- impute(activityData$steps, fun=mean)
```

##### 4. Make a histogram of the total number of steps taken each day 
```{r}
totalStepsFilled <- tapply(activityDataFilled$steps, activityDataFilled$date, sum)
qplot(totalStepsFilled, main = "Total Number of Steps Taken Each Day (No Missing Data)", xlab='total number of steps', binwidth=1000)
```

##### ... and Calculate and report the mean and median total number of steps taken per day. 
```{r}
totalStepsFilledMean <- mean(totalStepsFilled )
totalStepsFilledMedian <- median(totalStepsFilled )
```

----

## Are there differences in activity patterns between weekdays and weekends?
##### 1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
dayType <- function(date) {
    day <- weekdays(date)
    if (day %in% c("lunedì", "martedì", "mercoledì", "giovedì", "venerdì"))
        return("weekday")
    else if (day %in% c("sabato", "domenica"))
        return("weekend")
    else
        stop("invalid date")
}

activityDataFilled$date <- as.Date(activityDataFilled$date)
activityDataFilled$day <- sapply(activityDataFilled$date, FUN=dayType)
```

##### 2. Make a panel plot containing a time series plot

```{r}
avgActivityDataFilled <- aggregate(steps ~ interval + day, data=activityDataFilled, mean)
ggplot(avgActivityDataFilled, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("average number of steps")
```

